Optimizing Structural Properties Of Neural Networks With Genetic Algorithms

### 2015-08-07 paper/paper.tex
second final version

\author{T. Staudt, E. Schultheis}
\email{thomas.staudt@stud.uni-goettingen.de, erik.schultheis@uni-goettingen.de}
\affiliation{University of Göttingen}

\date{today}

\begin{abstract}
    In this article we examine how structural properties of neural
    networks, like the local synapse densities, affect their
    learning success for various tasks. To do so, we look at a rate based
    model for neural networks and apply a learning algorithm called FORCE to let the network learn 
		to reproduce either low or high frequent sinusoidals. A sophisticated matching algorithm
    allows us to quantify the learning success of a network so that the
    fitness of structural characteristics can be evaluated. We then use evolutionary
    optimization methods on these parameters, and found that for high frequencies, the
		$\feedback$ parameter can be increased  to improve learning results, at the cost of strongly 
		decreasing the ability to memorize low frequencies. Alternatively, using ring topology instead
		of random connections has a similar effect, but still retains the ability to learn lower frequencies.
\end{abstract}

@@

We chose the same rate based network architecture that is used in the
original FORCE publication \citep{FORCE} (therein denoted as architecture
A) and also in \citep{RM}, where our notation mainly stems from. So we look
at a network $\mathcal{N}$ with $N$ internal neurons and states $x_i$ for
$1 \le i \le N$ that loosely represent the neurons' membrane potential. The
firing rate $r_i$ of the $i$-th internal neuron is given by $r_i = \tanh
x_i$.  Furthermore, the network contains a readout neuron with the state
\begin{equation} 
z = \sum_{j=1}^{N}\omega\Ix{read}_{j}\,r_j
\end{equation}
where $\omega\Ix{read}$ denotes the readout weight vector.
Taking into account both internal dynamics as well as an external feedback
pathway, the dynamics $t \mapsto x_i(t)$ of the single neurons are governed
by \footnote{~Note that we did not include external input signals in this
formula, as is done in \citep{FORCE} and \citep{RM}.}
\begin{equation}
    \dot{x}_i(t) = -x_i(t) + \sum_{i=1}^N \omega\Ix{rec}_{ij}\, r_j(t) + 
                   \sum_{i=1}^R \omega\Ix{fb}_{ij}\, z_j(t)~,
\end{equation}
where we introduce the internal recurrent synapse weights
$\omega\Ix{rec}$ and the feedback synapse weights
$\omega\Ix{fb}$. 
We solved this system of differential equations by
Euler integration with a time steps $\mathrm{d}t = 0.01$, which was
also used in \citep{FORCE}. 
%(How we constructed $\omega\Ix{rec}$ and
%$\omega\Ix{fb}$ will be described in \emph{Parametric Random Networks})

\paragraph{Learning} Learning took place by stepwisely adapting the weights
$\omega\Ix{read}$ in such a way that the readout dynamic $t\mapsto z(t)$
should eventually match a periodic target pattern $t\mapsto
z'(t)$, henceforth called a \emph{task}, as accurately as possible. To
update $\omega\Ix{read}$ we chose variant A of the FORCE algorithm
\citep{FORCE}, which rapidly modifies the feedback loop to keep the errors
$|z(t) - z'(t)|$ small from the beginning on.

While the authors of \citep{FORCE} update the weights $\lambda\Ix{read}$ in
fixed intervals, we used an adaptive mechanism to determine how often
learning should occur. 
%This was initially introduced
%for performance reasons, as the single learning steps proved to be the main
%computational bottleneck, but we also found that the long time network
%dynamics tended to be slightly more accurate for adaptive learning steps.

\paragraph{Network Genotypes} 
In order to find out which structural elements of networks are important
for learning sucess, we introduce \emph{network genotypes}.
%Since we are not interested in single network
%properties, but rather want to find out which structural elements of
%networks are important for learning success, we introduced  \emph{network
%genotypes}.
%Such a genotype $G$ consists of a set $\Theta$ of parameters that determine
%which concrete networks are created when $G$ is expressed in a single
%phenotype. 
Technically, a genotype $G$ with parameters $\Theta$ may be understood as
a stochastic network generator that returns a network
$\mathcal{N}=G(\lambda)$ for every seed value $\lambda$.

The genotypes' parameters $\theta\in\Theta$ are either integer or floating
point values and can be classified as affecting (1) the topology of
$\omega\Ix{rec}$ or (2) the weight values of $\omega\Ix{rec}$ and
$\omega\Ix{fb}$. Optimized parameters of type (1) are the
occupation probability $\ERprob$ for sparse random topologies (Erdös-Renyi
topology), the neighborhood range $\RINGk$ for ring topologies, and a ratio
parameter $\ERtoRING$ used to interpolate between Erdös-Renyi and ring
topologies. Optimized parameters of type (2) are the feedback strength
$\feedback\propto \omega\Ix{fb}$ and the gain value
$\gain\propto\omega\Ix{rec}$.

\paragraph{Fitness} The next step was to define the \emph{fitness} of
a given genotype when confronted with a range of different tasks, e.g.
waves of different frequencies. To express this formally, we introduce
\emph{challenges}: For a given seed value $\lambda$, the challenge $C$
yields a task $z' = C(\lambda)$. If we now apply a success function
$S(\mathcal{N}, z')\in\{0, 1\}$ that decides whether the network $\mathcal{N}$
has successfully reproduced the desired output $z'$, the fitness $F_C(G)$
of a genotype $G$ for $C$ may be defined as the expectation value of
$\lambda \mapsto S(G(\lambda), C(\lambda))$. For fixed challenges, the
main problem of optimizing the genotype thus is to find parameters
$\Theta$ that maximize $F_C$.

@@

Our concrete goal for this article was to examine which choices of genotype
parameters are optimal for learning sinusoidal waves with relatively high
or low frequencies. 
Based on preliminary simulation results we chose
$\nu\in[0.3, 3]$ as frequency range for the challenge $C\ix{high}$, producing
high frequency waves, and $\nu\in[0.004, 0.04]$ for $C\ix{low}$, producing low
frequency waves. 
Optimizing the fitness functions $F_{C\ix{high}}$ or $F_{C\ix{low}}$ thus
results in genotypes with the desired frequency properties.

@@

\paragraph{Success}
To actually compute the fitness, we need a function $S$ that decides
whether a network succeeded to solve a task or not. To this end,
a measure $f(z', z)$ for the concordance of the target function $t\to z'(t)$ and
the network readout $t\to z(t)$ is sought. 
Though one could use a simple difference based metric, this has the
drawback that plain phase mismatches could produce high
differences for properly matching shapes.

@@

A measure that is time-shift independent and maximized for functions of
identical shape is the maximum of the cross correlation 
\begin{align}
    f\ix{cor}(z, z') = \max_t \frac{\left<z, z'(\cdot
    - t)\right>}{\sqrt{\left< z, z\right> \left<z', z' \right>}}~.
\end{align}
Algorithmically, the signals $z$ and $z'$ are split into overlapping chunks
which are correlated separately.  This saves computational power and
prevents minute differences in frequency to accumulate into a huge phase
shift, so the phase only has to remain relatively constant over the
timescale of a single chunk (which is about 10 seconds of simulated time).
%This measure becomes problematic for functions that are very close to zero
%for a longer time (order of chunk size), but since we are going to work
%with superpositions of sinusoidals, this drawback
%is unproblematic.

Using this measure of quality, the success function for a network
$\mathcal{N}$ with readout $z$ for the task $z'$ was (by empirical means)
taken to be 
\begin{equation}
    S(\mathcal{N}, z') = \begin{cases} 1,\quad\mathrm{if}~f\ix{cor}(z, z') \ge 0.95,\\
                                       0,\quad\mathrm{if}~f\ix{cor}(z, z') < 0.95~.
                         \end{cases}
\end{equation}
%The threshold value of 0.95 was determined empirically, as signals $r$ with
%$f(z, z') \ge 0.95$ met our assessment of what we considered a good
%reproduction of the target pattern.

\paragraph{Genetic Optimization}
Due to the stochastic nature of a genotype $G$'s fitness $F_C(G)$,
a very high number of samples would be required to numerically obtain sufficiently
smooth approximations of $F_C(G)$ for typical optimization methods like
gradient descend or simulated annealing. But as the computational cost for
even one learning process is considerably high, these methods were not
feasible.

@@

Therefore, we optimized $F_C$ using a genetic algorithm, which is inherently
capable of coping with the stochastic fitness function. In each iteration
of the algorithm, a generation $\mathcal{G}_t = \left\{ G_i~|~1\le i\le M\right\}$
of $M$ individuals (genotypes) is converted into a new one, $\mathcal{G}_{t+1}$,
according to the following scheme:
\begin{description}[itemsep=0ex, leftmargin=0pt, format=\normalfont]
 \item[1. Evaluation] The fitness $F_i = F_C(G_i)$ of each individual is
     measured using at least 50 samples (and adaptively more if the
     $(F_i)_{1\le i\le N}$ are spaced closely, to at most 100).
 \item[2. Selection] We choose $50\%$ of the most successful members of $G_t$ as survivors.
     With a small chance, a weaker generator can survive, in order to keep the gene pool more diverse.
\item[3. Reproduction] We then replace the $(1-p)|G_t|$ elements that did not make
    it into the next generation by either \emph{mutating} or \emph{recombining} the 
    survivors and get $\mathcal{G}_{t+1}$. This means we either change
    a single parameter randomly or take two individuals and randomly choose
    for each parameter from which parent to take the value.   
\end{description}

@@

\section{Results}
\subsection{Optimizing Genotypes For Frequency Ranges}
Since we wanted to examine how the ideal genotypes vary for different
types of problems, we examined how the choice of a frequency range affects
the optimized genotype parameters.
To this end we looked at sparse random networks (Erdös-Renyi topology with
connection probability $\ERprob$) and conducted optimizations of the
parameters $\gain$, $\feedback$ and $\ERprob$ for both high ($C\ix{high}$)
and low ($C\ix{low}$) frequency challenges (see FIG.
\ref{fig:freq_optimization}).

@@

The most striking result is the strong growth of the feedback for
$C\ix{high}$ to values higher than 14, while the value for $C\ix{low}$
converges quickly to approximately 1 (FIG. \ref{fig:freq_optimization} (b)).
This behavior may indicate that the feedback pathway must
overshadow the (maybe too slow) internal mechanics for high frequencies,
while it is the other way round for low frequencies.
Indeed, FIG. \ref{fig:freq_optimization} (a) shows that genotypes with
a lower gain value (on average) are more successful for high frequencies,
which seems to fit in: excessive recurrent dynamics with a long time
scale (compared to the period to be learned) may disturb the learning.
%That the single genotypes remain quite scattered throughout the whole
%optimization process, though, implies that the exact gain value is not
%critical for the success of a genotype.

@@

But how well was the fitness actually improved during the optimization?
When looking at FIG. \ref{fig:freq_optimization} (d), one sees that the
fitness for $C\ix{high}$ increased during nearly the whole process, while
the algorithm failed to further optimize the genotypes' fitness for
$C\ix{low}$ after about 20 generations. A major reason for this is
probably, that the initial population for the $C\ix{low}$ optimization
already contained genotypes with parameters similar to the (at least
locally) optimal ones.

One can also look at the success of the optimization algorithm in a more
specific way: FIG. \ref{fig:freqrange} shows how well an unoptimized
genotype $G_0$ handles fixed-frequency challenges when compared to the
genotypes $G\ix{high}$ and $G\ix{low}$ optimized for
$C\ix{high}$ and $C\ix{low}$. Here it becomes clear that $G\ix{high}$ can
indeed succeed at higher frequencies than $G_0$, but that it also fails at
learning lower frequencies -- which was expected. On the contrary, while
the high frequency capacities of $G\ix{low}$ have decreased when compared
to $G_0$, the low frequency capacities have only marginally improved. This
reinforces the impression that the optimization was unable to produce
genotypes that can cope with low frequencies.

@@

\subsection{Topology Optimizations}
To find out whether the topology may be used to improve the learning for
high frequencies without increasing the feedback, we performed an
optimization in which the network topologies were an interpolation between
Erdös-Renyi and ring topology.
%the $\RINGk$ nearest neighbors are connected (for non integer $\RINGk$, the
%fractional part determines the probability to use $[\RINGk]$ or $[\RINGk]+1$).
For this simulation, the parameters $\gain = 1.5$ and $\feedback = 2$ are
kept constant during the optimization process, leaving the connection
probability for Erdös-Renyi topology $\ERprob$, the ring connection length $\RINGk$
and the interpolation value $\ERtoRING$ between Erdös-Renyi and ring as
free parameters.

The results of the optimization are depicted in FIG.
\ref{fig:top_optimization}. Unfortunately, the improvement in fitness after the
first few generations is extremely low. The development of the ration
parameter, however, unambiguously shows that the selection process prefers
generators with a high fraction of ring topology and with $\RINGk
\approx 1.5$. The Erdös-Renyi connection probability $\ERprob$ shows a very
broad distribution, which is due to the weak dependence of the generated
networks on $\ERprob$ when $\ERtoRING$ approaches zero.

@@

While we only presented results concerning simple sinusoidals of varying
frequencies here, the code developed during our project provides
a framework which can be used to perform a variety of similar and extended
investigations; for example the inclusion of feed forward topologies and
arbitrary (higher dimensional) tasks with optional external input pathways
(e.g.\ as a timer signal to make the network synchronize better).

One peculiarity to be recognized throughout our project work is that the
optimization algorithm does not excel at providing strictly converging,
single most optimal parameter values, but that we are rather given more or
less broad ``survival ranges'' for each parameter as result. And though the
final genotypes varied quite a bit in their fitness, we did not observe
strong correlations between the different parameters in these ranges when
doing sporadic tests.

Indeed, looking at all of our simulations in their entirety, the genotype
parameters showed less influence on the learning results than we initially
expected, and during our inquiry it often seamed that the statistics of the
challenges might overshadow the effect of the parameters -- at least when
looking at more complex tasks than monofrequent waves. A wise improvement
would therefore be to make the challenges more deterministic, e.g.\ to use
the same tasks for all genotypes in each generation instead of drawing them
independently from some (wide range) distribution.

@@

Nevertheless we were able to produce successful optimization results
and could indeed show that adaption to high frequencies can not only be
accomplished by an implausible effect, namely the 
high feedback values, but also by changing the topological structure of the
network. In this regard it would certainly be interesting to further
inquire, how genotypes that allow for a more fine grained control over the
topology behave under the optimization.

>>>>

\author{T. Staudt, E. Schultheis}
\email{thomas.staudt@stud.uni-goettingen.de, erik.schultheis@stud.uni-goettingen.de}
\affiliation{University of Göttingen}

\date{\today}

\begin{abstract}
In this article we examine how structural properties of neural networks,
like the local synapse density, affect their learning success for various
tasks. To do so, we look at a rate based model for neural networks and
apply a learning algorithm called FORCE to teach the network to
reproduce either low or high frequent sinusoidals. A sophisticated matching
algorithm allows us to quantify the learning success of the networks so that
the fitness of structural characteristics (expressed via parameters) can be
evaluated. We then use evolutionary optimization methods on these
parameters, 
and find that the feedback pathway may be strengthened in order
to improve the learning results for high frequencies but at the cost of
decreasing the ability to memorize low frequencies.
%and find that for high frequencies, the feedback pathway
%can be strengthened to improve learning results at the cost of decreasing
%the ability to memorize low frequencies. 
Alternatively, using a ring topology model instead of random connections
has a similar effect, yet still retains the ability to learn lower
frequencies better.
\end{abstract}

@@

We chose the same rate based network architecture that is used in the
original FORCE publication \citep{FORCE} (therein denoted as architecture
A) and also in \citep{RM}, where our notation mainly stems from. So we look
at a network $\mathcal{N}$ with $N$ internal neurons and states $x_i$ for
$1 \le i \le N$ that loosely represent the neurons' membrane potential. The
firing rate $r_i$ of the $i$-th internal neuron is given by $r_i = \tanh
x_i$.  Furthermore, the network contains a readout neuron with the state
\begin{equation} 
z = \sum_{j=1}^{N}\omega\Ix{read}_{j}\,r_j~,
\end{equation}
where $\omega\Ix{read}$ denotes the readout weight vector.
Taking into account both internal dynamics as well as an external feedback
pathway, the dynamics $t \mapsto x_i(t)$ of the single neurons are governed
by \footnote{Note that we did not include external input signals in this
formula, as is done in e.g.\ \citep{RM}.}
\begin{equation}
    \dot{x}_i(t) = -x_i(t) + \sum_{i=1}^N \omega\Ix{rec}_{ij}\, r_j(t) + 
                   \sum_{i=1}^R \omega\Ix{fb}_{ij}\, z_j(t)~,
\end{equation}
where we introduce the internal recurrent synapse weights
$\omega\Ix{rec}$ and the feedback synapse weights
$\omega\Ix{fb}$. 
We solved this system of differential equations by
Euler integration with a time steps $\mathrm{d}t = 0.01$, which was
also used in \citep{FORCE}. 
%(How we constructed $\omega\Ix{rec}$ and
%$\omega\Ix{fb}$ will be described in \emph{Parametric Random Networks})

\paragraph{Learning} Learning took place by stepwisely adapting the weights
$\omega\Ix{read}$ in such a way that the readout dynamic $t\mapsto z(t)$
should eventually match a periodic target pattern $t\mapsto
z'(t)$, henceforth called a \emph{task}, as accurately as possible. To
update $\omega\Ix{read}$ we chose variant A of the FORCE algorithm
\citep{FORCE}, which rapidly modifies the feedback loop to keep the errors
$|z(t) - z'(t)|$ small from the beginning on.
While the authors of \citep{FORCE} update the weights $\omega\Ix{read}$ in
fixed intervals, we use an adaptive mechanism to determine how often
learning should occur. 
%This was initially introduced
%for performance reasons, as the single learning steps proved to be the main
%computational bottleneck, but we also found that the long time network
%dynamics tended to be slightly more accurate for adaptive learning steps.

\paragraph{Network Genotypes} 
In order to find out which structural elements of networks are important
for learning success, we introduce \emph{network genotypes}.
%Since we are not interested in single network
%properties, but rather want to find out which structural elements of
%networks are important for learning success, we introduced  \emph{network
%genotypes}.
%Such a genotype $G$ consists of a set $\Theta$ of parameters that determine
%which concrete networks are created when $G$ is expressed in a single
%phenotype. 
Technically, a genotype $G$ with parameters $\Theta$ may be understood as
a stochastic network generator that returns a network
$\mathcal{N}=G(\lambda)$ for every seed value $\lambda$.

The genotypes' parameters $\theta\in\Theta$ are either integer or floating
point values and can be classified as affecting (1) the topology of
$\omega\Ix{rec}$ or (2) the weight values of $\omega\Ix{rec}$ and
$\omega\Ix{fb}$. Optimizable parameters of type (1) are the
occupation probability $\ERprob$ for sparse random topologies (Erdös-Renyi
topology), the neighborhood range $\RINGk$ for ring topologies (where
either $\lfloor k\rfloor$ or, with probability $k - \lfloor k\rfloor$,
$\lfloor k\rfloor+1$ nearest neighbors are connected),
%fractional part determines the probability to use $[\RINGk]$ or $[\RINGk]+1$).
and an interpolation parameter $\ERtoRING$ used to interpolate between Erdös-Renyi
and ring topology. Optimizable parameters of type (2) are the feedback
strength $\feedback\propto \omega\Ix{fb}$ and the gain value
$\gain\propto\omega\Ix{rec}$.

\paragraph{Fitness} The next step is to define the \emph{fitness} of
a given genotype when confronted with a range of different tasks, e.g.
waves of different frequencies. To express this formally, we introduce
\emph{challenges}: For a given seed value $\lambda$, the challenge $C$
yields a task $z' = C(\lambda)$. If we now apply a success function
$S(\mathcal{N}, z')\in\{0, 1\}$ that decides whether the network $\mathcal{N}$
has successfully reproduced the desired output $z'$, the fitness $F_C(G)$
of a genotype $G$ for $C$ may be defined as the expectation value of
$\lambda \mapsto S(G(\lambda), C(\lambda))$. For fixed challenges, the
main problem of optimizing the genotype thus is to find parameters
$\Theta$ that maximize $F_C$.

@@

Our concrete goal for this article was to examine which choices of genotype
parameters are optimal for learning sinusoidal waves with relatively high
or low frequencies. 
Based on preliminary simulation results we chose $\nu\in[0.3, 3]$ as
frequency range for the challenge $C\ix{high}$, producing high frequency
tasks, and $\nu\in[0.004, 0.04]$ for $C\ix{low}$, producing low frequency
tasks. 
Optimizing the fitness functions $F_{C\ix{high}}$ or $F_{C\ix{low}}$ will
thus yield genotypes with the desired frequency properties.

@@

\paragraph{Success}
To actually compute the fitness, we need a function $S$ that decides
whether a network succeeded to solve a task or not. To this end,
a measure $f(z, z')$ for the concordance of the target function $t\to z'(t)$ and
the network readout $t\to z(t)$ is sought. 
Though one could use a simple difference based metric, this has the
drawback that plain phase mismatches could produce high
differences for properly matching shapes.

@@

A measure that is timeshift independent and maximized for functions of
identical shape is the maximum of the cross correlation 
\begin{align}
    f\ix{cor}(z, z') = \max_t \frac{\left<z, z'(\cdot
    - t)\right>}{\sqrt{\left< z, z\right> \left<z', z' \right>}}~.
\end{align}
Algorithmically, the signals $z$ and $z'$ are split into overlapping chunks
which are correlated separately.  This saves computational power and
prevents minute differences in frequency to accumulate into a huge phase
shift, so the phase only has to remain relatively constant over the
timescale of a single chunk (which is about 10 seconds of simulated time).
%This measure becomes problematic for functions that are very close to zero
%for a longer time (order of chunk size), but since we are going to work
%with superpositions of sinusoidals, this drawback
%is unproblematic.

Using this measure of quality, the success function for a network
$\mathcal{N}$ with readout $z$ for the task $z'$ is (by empirical means)
taken to be 
\begin{equation}
    S(\mathcal{N}, z') = \begin{cases} 1,\quad\mathrm{if}~f\ix{cor}(z, z') \ge 0.95,\\
                                       0,\quad\mathrm{if}~f\ix{cor}(z, z') < 0.95~.
                         \end{cases}
\end{equation}
%The threshold value of 0.95 was determined empirically, as signals $r$ with
%$f(z, z') \ge 0.95$ met our assessment of what we considered a good
%reproduction of the target pattern.

\paragraph{Genetic Optimization}
Due to the stochastic nature of a genotype's fitness $F_C(G)$,
a very high number of samples would be required to numerically obtain sufficiently
smooth approximations of $F_C(G)$ for typical optimization methods like
gradient descend or simulated annealing. But as the computational cost for
even one learning process is considerably high, these methods are not
feasible.

@@

Therefore, we optimize $F_C$ using a genetic algorithm, which is inherently
capable of coping with the stochastic fitness function. In each iteration
of the algorithm, a generation $\mathcal{G}_t = \left\{ G_i~|~1\le i\le M\right\}$
of $M$ individuals (genotypes) is converted into a new one, $\mathcal{G}_{t+1}$,
according to the following scheme:
\begin{description}[itemsep=0ex, leftmargin=0pt, format=\normalfont]
 \item[1. Evaluation] The fitness $F_i = F_C(G_i)$ of each individual is
     measured using at least 50 samples (and adaptively more if the 
     $(F_i)_{1\le i\le N}$ are tightly spaced, to at most 100).
 \item[2. Selection] We choose $50\%$ of the most successful members of $G_t$ as survivors.
     With a small chance, a weaker generator can survive, in order to keep the gene pool more diverse.
\item[3. Reproduction] We then replace the $|G_t|/2$ elements that did not make
    it for the next generation by either \emph{mutating} or \emph{recombining} the 
    survivors and get $\mathcal{G}_{t+1}$. This means we either change
    a single parameter randomly or take two individuals and randomly choose
    for every parameter from which parent to take the value from.
\end{description}

@@

\section{Results}
\subsection{Optimizing Genotypes For High And Low Frequencies}
Since we want to analyze how the ideal genotypes vary for different types
of problems, we examine how the choice of either low or high frequencies
affects the optimized genotype parameters.
We first look at networks with Erdös-Renyi topology, for which we conduct
optimizations of the parameters $\gain$, $\feedback$ and $\ERprob$ for both
high ($C\ix{high}$) and low ($C\ix{low}$) frequency challenges (see FIG.
\ref{fig:freq_optimization}).

@@

The most striking result is the strong growth of the feedback for
$C\ix{high}$ to values higher than 14, while the value for $C\ix{low}$
converges quickly to approximately 1 (FIG. \ref{fig:freq_optimization} (b)).
This behavior may indicate that the feedback pathway must
overshadow the (maybe too slow) internal mechanics for high frequencies,
while there is a contrary effect for low frequencies.
Indeed, FIG. \ref{fig:freq_optimization} (a) shows that genotypes with
a lower gain value (on average) are more successful for high frequencies,
which seems to fit in: excessive recurrent dynamics with a long time
scale (compared to the period to be learned) may disturb the learning.
%That the single genotypes remain quite scattered throughout the whole
%optimization process, though, implies that the exact gain value is not
%critical for the success of a genotype.

@@

But how much is the fitness actually improved during the optimization?
When looking at FIG. \ref{fig:freq_optimization} (d), one sees that the
fitness for $C\ix{high}$ increases during nearly the whole process, while
the algorithm fails to further optimize the genotypes' fitness for
$C\ix{low}$ after about 20 generations. A major reason for this is
probably, that the initial population for the $C\ix{low}$ optimization
already contains genotypes with parameters similar to the (at least
locally) optimal ones.

One can also look at the success of the optimization algorithm in a more
specific way: FIG. \ref{fig:freqrange} shows how well an unoptimized
genotype $G_0$ handles fixed-frequency challenges when compared to the
genotypes $G\ix{high}$ and $G\ix{low}$ optimized for
$C\ix{high}$ and $C\ix{low}$. Here it becomes clear that $G\ix{high}$ can
indeed succeed at higher frequencies than $G_0$, but that it also fails at
learning lower frequencies -- which was expected. However, while
the high frequency capacities of $G\ix{low}$ have decreased when compared
to $G_0$, the low frequency capacities have only marginally improved. This
reinforces the impression that the optimization is unable to produce
genotypes that can cope with low frequencies significantly better than the
initial population.

@@

\subsection{Topology Optimizations}
To find out whether the topology may be used to improve the learning for
high frequencies without increasing the feedback, we perform an
optimization in which the network topology is an interpolation between
Erdös-Renyi and ring topology.
%the $\RINGk$ nearest neighbors are connected (for non integer $\RINGk$, the
%fractional part determines the probability to use $[\RINGk]$ or $[\RINGk]+1$).
For this simulation, the parameters $\gain = 1.5$ and $\feedback = 2$ are
kept constant during the optimization process, leaving the connection
probability $\ERprob$ for the Erdös-Renyi topology, the ring neighborhood range $\RINGk$
and the interpolation value $\ERtoRING$ between Erdös-Renyi and ring topology as
free parameters.

The results of the optimization are depicted in FIG.
\ref{fig:top_optimization}. Unfortunately, the improvement in fitness after the
first few generations is extremely low. The development of the interpolation
parameter $q$, however, unambiguously shows that the selection process prefers
generators with a high fraction of ring topology and with $\RINGk
\approx 1.5$. The Erdös-Renyi connection probability $\ERprob$ shows a very
broad distribution, which is due to the weak dependence of the generated
networks on $\ERprob$ when $\ERtoRING$ approaches zero.

@@

While we only presented results concerning simple sinusoidals of varying
frequencies in this article, the code developed during our project provides
a framework which can be used to perform a variety of similar and extended
investigations; for example the inclusion of feed forward topologies and
arbitrary (higher dimensional) tasks with optional external input pathways
(e.g.\ as a timer signal to make the network synchronize better).

One peculiarity to be recognized throughout our project work is that the
optimization algorithm does not excel at providing strictly converging,
single most optimal parameter values, but that we are rather given more or
less broad ``survival ranges'' for each parameter as result. And though the
final genotypes vary quite a bit in their fitness, we did not observe
strong correlations between the different parameters in these ranges when
doing sporadic tests.

Indeed, looking at all of our simulations in their entirety, the genotype
parameters show less influence on the learning results than we initially
expected, and during our inquiry it often seamed that the statistics of the
challenges might overshadow the effect of the parameters -- at least when
looking at more complex tasks than monofrequent waves. A wise improvement
would therefore be to make the challenges more deterministic, e.g.\ to use
the same tasks for all genotypes in each generation instead of drawing them
independently from some (wide range) distribution.

@@

Nevertheless we were able to produce successful optimization results
and could indeed show that adaption to high frequencies can not only be
accomplished by an implausible effect, namely the 
high feedback values, but also by changing the topological structure of the
network. In this regard it would certainly be interesting to further
inquire how genotypes that allow for a more fine grained control over the
topology behave under the optimization.
